### 2019-03-01 UnderstandabilityPaper/machinelearning.tex
Fix up understandability section

% STRUCTURE:

%  - Understandability in AI
%  - Theorem proving as example:
%   - Least informative is a yes/no/unknown answer
%   -

\documentclass{article}
\begin{document}
The problem of making machine-generated artefacts more understandable to human
users is quite clear in the field of \emph{automated theorem
  proving}~\cite{sutcliffe2001evaluating}: the task of deciding whether a given
formal statement follows from a given set of premeses.  The least informative
approaches, such algorithms based on the popular \emph{resolution}
rule~\cite{robinson1965machine}, produce nothing more than a ``yes'', ``no'' or
``unknown'' response. Not only is this devoid of \emph{explanation}, but it also
hides the effects of any bugs; requiring the user to either trust the results,
or verify the implementation.

@@

% A similar pattern has emerged when it comes to understanding

% I meant to send this earlier, in case it's useful. These are just some
% things that came to mind regarding "explainable" AI and machine
% learning. Lots of rambling, but hopefully some insight into current
% machine-learning thinking (as I understand it, as somewhat of an
% outsider!).

% I think a key idea is "differentiable programming", which captures the
% essence of how machine learning models like neural networks work, and
% precisely what makes them different from traditional computer programs
% (i.e. they're differentiable!), without ignoring the details ("black
% box") or using flimsy analogies ("brain-like"):

% https://fluxml.ai/2019/02/07/what-is-differentiable-programming.html

% When run on an input, a differentiable program doesn't just give an
% output, it also gives us the partial derivatives of that output
% w.r.t. each of the program's parameters. This tells us exactly how to
% adjust those parameters (via backpropagation) to reduce the output error
% for that particular input. Doing this enough times for enough inputs
% will hopefully find a general solution.

% A neural network is an example of a differentiable program, one which is
% just a big collection of identical expressions ("neurons") composed
% together. This is incredibly general (they are universal function
% approximators), but it doesn't have much "knowledge" or "structure"
% baked in: the behaviour of this program depends almost completely on the
% parameters, so everything it "learns" must come from input/output
% examples.

% The convolutional neural networks that have come to dominate image
% processing are also differentiable programs, but they contain more
% knowledge/structure. Instead of general-purpose 'neurons', these use
% 'convolutional layers' which are also expressions that get composed
% together. The difference is that these expressions hard-code the 2D grid
% structure that we expect to find in raster images. This makes them less
% general than neural networks, since they assume the data fits this 2D
% grid structure, but they also don't have to learn that structure from
% examples. Hence they need fewer examples, but we can also be more
% confident that the input/output relationships they learn involve 2D
% patches of image content (like shapes or textures), rather than some
% arbitrary numerical relationship between pixels.

% Taking things much further, there are differentiable programs which do
% ray tracing ( https://people.csail.mit.edu/tzumao/diffrt/ ). These can
% be used to learn the position of 3D objects in an image, the camera
% position, illumination, etc. Again, this is less general since it
% assumes the input is an image of a 3D scene, but again it doesn't have
% to learn this knowledge from examples. The learned parameters are easily
% "explained", as they are the usual rendering parameters of camera
% position, rotation, etc.

% Another example is learning to recognise and generate hand-written
% characters by using a differentiable program for drawing lines:
% https://web.mit.edu/cocosci/Papers/Science-2015-Lake-1332-8.pdf

% In this way, there are two tradeoffs being made:

%  - By hard-coding our knowledge of the domain into a program, like using
%    a ray-tracer for images of the 3D world, or pen-based drawing for
%    images of hand-written characters, we don't need as much training
%    data for the system to 'learn' what's going on. Unfortunately, we end
%    up with less general systems, e.g. we might apply the same
%    neural-network architecture to all sorts of problems (with enough
%    data), but we can't apply the ray-tracer to e.g. audio data.

%  - This same tradeoff also has implications for understandability and
%    explainability: the parameters of a ray-tracer are directly
%    understandable, based on how they're used in the program, e.g. "how
%    bright is this light source" or "how big is this object". In
%    contrast, the parameters of something like a neural network make
%    little sense on their own, they just mean "how important is this
%    output for that neuron?".

% STUFF BELOW IS MORE TANGENTIAL

% There are probably parallels to be drawn w.r.t. feature engineering
% too. "Feature extraction" turns raw data into numbers for the learning
% algorithm. This used to be done by hand, so we knew what the numbers
% meant (e.g. average brightness, number of edges, etc. this is what
% Katya's 'recurrent clustering' algorithm does for symbolic expressions).
% With "deep nets", such features are figured out automatically, and we
% don't know what they mean. Hence we gain generality (automatically
% discovered features tend to out-perform hand-engineered ones), but we
% lose understanding/explainability, and we become more dependent on
% training data.

% Things like "deep dream" can try to demonstrate what these features
% seem to represent, but not much more. As far as I understand it, deep
% dream works by picking a particular neuron/feature, providing an
% arbitrary input image to the system, and iteratively tweaking that image
% so that feature increases. The "explanation" of that feature is the
% changes that were made to the image, e.g. if the changes look
% eye-shaped, then we say that feature is for detecting eyes. This is, of
% course, open to interpretation.

% Interestingly, exactly the same technique is used for generating
% adversarial inputs, where an input (usually to a classifier) is tweaked
% until it gets classified incorrectly:

% % https://en.wikipedia.org/wiki/Adversarial_machine_learning

% Adversarial inputs show how limited this sort of "explanation" is, since
% these changes might look like random noise:
% https://blog.openai.com/adversarial-example-research/

% Or might even just be a single pixel:
% https://arxiv.org/abs/1710.08864

% Another nice example of "explaining" a machine learning model by
% demonstrating what its features represent is the variational
% autoencoder.

% Normal autoencoders 'compress' inputs down to a sequence of numbers
% representing their important features. Unfortunately these are hard to
% understand in isolation (sometimes called "holographic reduced
% representations", all of the components depend on all of the others).

% Variational autoencoders compress down to a multi-dimensional gaussian
% distribution instead. By randomly sampling from this gaussian, the
% representation is forced to be robust against small changes, and this
% imposes a meaningful structure on the space; such that adjusting the
% centre of the gaussian will alter the resulting data, e.g. by making a
% person's hair darker; or changing their gender; or giving them glasses;
% etc.

% https://towardsdatascience.com/intuitively-understanding-variational-autoencoders-1bfe67eb5daf
% (Especially the "Vector arithmetic" section)

% Although this property can't actually be relied upon!
% http://paulrubenstein.co.uk/variational-autoencoders-are-not-autoencoders/

% A similar effect to the structured "latent space" of the VAE has been
% found with word2vec:

% % https://en.wikipedia.org/wiki/Word2vec#Preservation_of_semantic_and_syntactic_relationships

% word2vec turns words into vectors of numbers, based on co-occurrences of
% words in some training data. These vectors turn out to have some
% "meaning" in the sense that vector arithmetic can make semantic changes.
% For example if we do the sum:

%   vector("king") - vector("man") + vector("woman")

% The resulting vector is very close to vector("queen").

% These sorts of machine learning system are unsupervised, so they don't
% solve any particular problem. However, the representations they learn
% can be useful for other task-specific systems.

>>>>

\documentclass{article}
\usepackage{cite}
\begin{document}
The problem of making machine-generated artefacts more understandable to human
users is quite clear in the field of \emph{automated theorem
  proving}~\cite{sutcliffe2001evaluating}: the task of deciding whether a given
formal statement follows from a given set of premeses.  The least informative
approaches, such algorithms based on the popular \emph{resolution}
rule~\cite{robinson1965machine}, produce nothing more than a ``yes'', ``no'' or
``unknown'' response. Not only is this devoid of \emph{explanation}, but it also
hides the effects of any bugs; requiring the user to either trust the results,
or verify the implementation.

@@

A similar problem of understandability has emerged in the field of \emph{machine
  learning} (ML), where statistical methods are used to make predictions or
decisions about data, often without human
involvement~\cite{alpaydin2014introduction}. In particular there is a tradeoff
between the generality of an approach and how easily its resulting behaviour can
be understood.

Many ML approaches can be characterised as constructing a computer program (or
``model'') consisting of two parts: an overall structure or \emph{architecture},
which remains fixed; and a set of adjustable \emph{parameters}, which are
inferred or ``learned'' from data (e.g. \emph{training data} of desired
input/output examples). One particularly simple architecture is the
\emph{decision tree}: nested boolean queries of the input, often used for
\emph{classification}~\cite{safavian1991survey}. These queries are parameters,
and are chosen based on how efficiently they separate the classes given in the
training data.  Decision trees are nominally understandable, since their
behaviour on a given input traces a single path through these queries; yet they
perform poorly compared to other ML algorithms. The \emph{random forest}
approach yields better results by combining many decision trees and having them
vote on the overall outcome~\cite{randomforests}, although such ensemble
behaviour is more difficult to reason about than that of a single tree.

Recent research has focused on highly expressive classes of models such as
\emph{differentiable programming}~\cite{wang2018demystifying}, whose
architectures output not only a (numerical) answer, but also partial derivatives
with respect to the parameters; and \emph{probabilistic programming}, which
samples from a \emph{distribution} conditioned on the training data. Both
frameworks allow arbitrary architectures, specified via Turing-complete
languages, and provide efficient, composable methods for optimising the
parameters (e.g. Stochastic Gradient Descent and Markov Chain Monte Carlo,
respectively) to minimise arbitrary loss functions (e.g. the squared output
error on a training data set).

With such expressive formalisms, the conflict between the generality of a model
and its understandability becomes clear. Task-specific architectures require
fewer parameters than general-purpose approaches and perform well with little
training data, for example hand-written characters can be classified based on a
single example if we allow our model to assume the given images are generated by
pen strokes~\cite{lake2015human}. Likewise we can infer parameters of a 3D scene
(such as object position and lighting) from images if we embed a ray-tracer into
our model~\cite{li2018differentiable}; embedding a physics engine can provide
predictions about these scenes, which are useful e.g. for robot
controllers~\cite{degrave2016differentiable}.

The parameters of such programs are easily understood: they describe the pen
strokes, geometric and physical properties of the simulated system, inferred
from the given images. However, such specificity makes these implementations
unsuitable for anything else. The choice of such high-level, task-specific
components is performed by the user, and encodes some of their domain knowledge
into the structure of the solution, such that it doesn't have to be learned.
This is similar to how logics like HOL can directly represent high-level
concepts (e.g. natural numbers and sets) but whose proof methods have limited
reusability due to the difficulty of automating such high-level reasoning.

At the other end of the spectrum we have differentiable programs with general
purpose architectures, like \emph{neural networks} (capable of universal
function approximation~\cite{funahashi1989approximate}). These are compositions
of a large number of identical sub-expressions (``neurons''), whose parameters
(``weights'') scale the input values, and hence the contribution of each
sub-expression to the whole. Such architectures encode essentially no domain
knowledge, requiring much larger training sets than task-specific algorithms in
order to ``learn'' these details.  The behaviour of general purpose models
depends so heavily on their (many) parameters, that understanding their
behaviour becomes difficult and they are often treated as inscrutable ``black
boxes''; akin to the large (un)SAT proof described above.

Full generality is seldom required: even if we don't want to constrain a model's
internal operation, the input and output data formats will always be known. Such
knowledge can be encoded in the architecture through the use of
\emph{convolutional layers}, for example forcing a 2D grid structure when
dealing with images~\cite{krizhevsky2012imagenet}. Introducing task-specific
components like this can aid in understanding the resulting models, since they
are constrained to operate on 2D patches of image content (like shapes or
textures), rather than potentially arbitrary numerical relationships between
pixels.

Understanding exactly how these programs make their decisions is an active area
of research, known as \emph{explainable artificial intelligence}
(XAI)~\cite{dovsilovic2018explainable,doshi2017towards,molnar2018interpretable},
with popular approaches such as saliency maps~\cite{simonyan2013deep} tending to
reverse-engineer the factors which lead to a model's decisions (judging the
saliency of each pixel based on how strongly it would effect the output
prediction if adjusted). These methods appear intuitive, e.g. producing
visualisations highlighting a particular object in a scene as the reason for its
classification; akin to the proof objects emitted by theorem provers. Yet this
can obscure the difficulty of interpreting such high-dimensional decision
boundaries. In particular, these explanations or justifications can be
fundamentally altered by imperceptible adjustments to the
input~\cite{ghorbani2017interpretation}. Similar adjustments can also change a
model's output, leading to the field of \emph{adversarial machine
  learning}~\cite{goodfellow2014explaining}; adjustments to even a single pixel
can not only cause a system to mislabel an input, but to give high confidence to
its erroneous result~\cite{su2019one}. Such fragility and opaqueness has
important implications as these techniques begin to have decision making
applications in life-critical domains such as automotive
control~\cite{bojarski2016end} and medical
diagnosis~\cite{esteva2017dermatologist}.
